---
title: "acoustic_to_camtrapdp"
author: 
- "Sanne Govaert"
- "Julia Wiel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(here)
library(dplyr)
library(tidyr)
library(frictionless)
library(camtrapdp)
library(lubridate)
library(readxl)
```

# Import data

```{r}
data_raw <- read_xlsx(here::here("datasets", "maryland", "data", "raw",
 "2018_UnivMD_Maryland_86896 DEP1.xlsx"), sheet = 'DATA TABLE FOR EXPORT')
selections_raw <- read.delim2(here::here("datasets", "maryland", "data", "raw",
 "Fin_merged.selections.txt")) %>% sample_n(500)
```

# Create resources

## Inspect data

```{r}
(data_raw_unique_values_per_column <-
  data_raw %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "column", values_to = "n_unique"))

(selections_unique_values_per_column <-
  selections %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "column", values_to = "n_unique"))
```

## Add information to dataset

# data
```{r}
# Parse deploymentStart and deploymentEnd in UTC
data <- data_raw %>%
  dplyr::mutate(
    # Clean deployTime and recoveryTime
    deployTime_clean = sub("^\\d{4}-\\d{2}-\\d{2} ", "", deployTime),
    recoveryTime_clean = sub("^\\d{4}-\\d{2}-\\d{2} ", "", recoveryTime),
    # Combine date and time for deployment start/end
    deploy_datetime_str = paste(deployDate, deployTime_clean),
    recovery_datetime_str = paste(recoveryDate, recoveryTime_clean)
  ) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    deploymentStart = format(lubridate::ymd_hms(deploy_datetime_str, tz = "UTC"), "%Y-%m-%dT%H:%M:%S%z", tz = "UTC"),
    deploymentStart = sub("(\\+\\d{2})(\\d{2})$", "\\1:\\2", deploymentStart),
    deploymentEnd = format(lubridate::ymd_hms(recovery_datetime_str, tz = "UTC"), "%Y-%m-%dT%H:%M:%S%z", tz = "UTC"),
    deploymentEnd = sub("(\\+\\d{2})(\\d{2})$", "\\1:\\2", deploymentEnd)
  ) %>%
  dplyr::ungroup()

```

# selections
```{r}
selections
## Create deploymentID from Begin.Path in selections

# Split Begin.Path by "/"
split_path <- strsplit(selections$Begin.Path, '\\\\')
# Extract elements by position (e.g., 5th, 6th, 7th)
deploymentID <- sapply(split_path, function(x) {
  # Split x[7] by "_"
  parts <- strsplit(x[7], "_")[[1]]
  # Remove the first term and paste the rest back together
  new_x7 <- paste(parts[-1], collapse = "_")
  new_x7 <- gsub("MD", "Dep", new_x7)
  # Combine x[6] and the modified x[7]
  paste(x[6], new_x7, sep = "_")
})
# Add to selections
selections$deploymentID <- deploymentID

## Timestamp of the selection
selections$timestamp <- sapply(sub("\\.aif$", "", selections$Begin.File), function(x) {
  parts <- strsplit(x, "_")[[1]]
  # Assume last two substrings are date and time
  date_str <- parts[length(parts)-1]
  time_str <- parts[length(parts)]
  # Try to parse date and time
  dt <- suppressWarnings(lubridate::ymd_hms(paste(date_str, time_str), tz = "UTC"))
  # Format as ISO 8601 with offset
  ts <- format(dt, "%Y-%m-%dT%H:%M:%S%z", tz = "UTC")
  ts <- sub("(\\+\\d{2})(\\d{2})$", "\\1:\\2", ts)
  return(ts)
})

## ObservationID
selections$observationID <- paste(sub("\\.aif$", "", selections$Begin.File), selections$Selection,
 sep = "_")

## scientificName from Tags
selections <- selections %>%
  dplyr::mutate(
    scientificName = dplyr::case_when(
      Tags == "1-4-f" ~ "Balaenoptera physalus",
      Tags == "1-2-f" ~ "Balaenoptera physalus",
      Tags == "1-1-f" ~ "Balaenoptera physalus",
      TRUE ~ NA_character_
    )
  )
```
## Create deployments resource

```{r}
deployments <-
  data %>% 
  dplyr::mutate(
    deploymentID = data$KEY_deployment_ID, # Required
    locationID = data$nominalID,
    locationName = NA_character_,
    latitude = data$latitudeDeployed_DecDeg, # Required
    longitude = data$longitudeDeployed_DecDeg, # Required
    coordinateUncertainty = NA_integer_,
    elevation = NA_real_,
    deploymentStart = data$deploymentStart, # Required
    deploymentEnd = data$deploymentEnd, # Required
    setupBy = NA_character_,
    deviceID = data$containerID,
    deviceModel = data$recorderDeviceType,
    devicePlatform = "mooring",
    deviceDelay = NA_integer_,
    deviceHeight = NA_real_,
    deviceDepth = data$deploymentDepth_Meters,
    deviceTilt = NA_integer_,
    deviceHeading = NA_integer_,
    recordingSchedule = "continuous",
    detectionDistance = NA_real_,
    baitUse = NA,
    locationType = NA_character_,
    habitat = NA_character_,
    deploymentGroups = data$arrayName,
    deploymentTags = NA_character_,
    deploymentComments = data$notes,
    .keep = "none"
  ) %>% 
  dplyr::arrange(deploymentID) %>%
  dplyr::select(
    deploymentID, locationID, locationName, latitude, longitude,
    coordinateUncertainty, elevation, deploymentStart, deploymentEnd, setupBy,
    deviceID, deviceModel, devicePlatform, deviceDelay, deviceHeight,
    deviceDepth, deviceTilt, deviceHeading, recordingSchedule,
    detectionDistance, baitUse, locationType, habitat, deploymentGroups,
    deploymentTags, deploymentComments
  )
```

## Create media resource

```{r}
media <-
  selections %>% 
  dplyr::mutate(
    mediaID = selections$Begin.File, # Required
    deploymentID = selections$deploymentID, # Required
    captureMethod = "schedule",
    timestamp = selections$timestamp, # Required 
    duration = NA_real_,
    filePath = selections$Begin.Path,# Required
    filePublic = FALSE, # Required
    fileName = NA_character_,
    fileMediatype = "audio/aiff", # Required
    exifData = NA,
    bitDepth = 16,
    samplingFrequency = 16000,
    gain = 23.5,
    channels = NA_integer_,
    favorite = NA,
    mediaComments = NA_character_,
    .keep = "none"
  ) %>% 
  dplyr::arrange(mediaID) %>%
  dplyr::distinct() %>% 
  dplyr::select(
    mediaID, deploymentID, captureMethod, timestamp, duration, filePath,
    filePublic, fileName, fileMediatype, exifData, bitDepth, samplingFrequency,
    gain, channels, duration, favorite, mediaComments
  )
```

## Create observations resource

```{r}
observations <-
  selections %>% 
  dplyr::mutate(
    observationID = selections$observationID, # Required
    deploymentID = selections$deploymentID, # Required
    mediaID = selections$Begin.File,
    eventID = NA_character_, # Necessary to use read_camtrapdp(), see https://github.com/inbo/camtrapdp/issues/162
    eventStart = selections$Begin.Time..s., # format(data$detected_time_parsed, "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"), # Required
    eventEnd = selections$End.Time..s., # Required
    observationLevel = "interval", # Required
    observationType = "animal", # Required 
    deviceSetupType = NA_character_,
    scientificName = selections$scientificName,  # Required
    count = NA_integer_,
    lifeStage = NA_character_,
    sex = NA_character_,
    behavior = NA_character_,
    individualID = NA_character_,
    individualPositionRadius = NA_real_,
    individualPositionAngle = NA_real_,
    individualSpeed = NA_real_,
    bboxX = NA_real_,
    bboxY = NA_real_,
    bboxWidth = NA_real_,
    bboxHeight = NA_real_,
    frequencyLow = NA_real_,
    frequencyHigh = NA_real_,
    classificationMethod = "machine",
    classifiedBy = NA_character_,
    classificationTimestamp = NA_character_,
    classificationProbability = selections$Score,
    classificationConfirmation = NA_character_,
    observationTags = NA_character_,
    observationComments = selections$Notes,
    .keep = "none"
  ) %>% 
  dplyr::arrange(observationID) %>% 
  dplyr::select(
    observationID, deploymentID, mediaID, eventID, eventStart,
    eventEnd, observationLevel, observationType, deviceSetupType,
    scientificName, count, lifeStage, sex, behavior, individualID,
    individualPositionRadius, individualPositionAngle, individualSpeed,
    bboxX, bboxY, bboxWidth, bboxHeight, frequencyLow, frequencyHigh,
    classificationMethod, classifiedBy, classificationTimestamp,
    classificationProbability, classificationConfirmation, observationTags,
    observationComments
  )
```

# Create data package

```{r}
acoustic <-
  frictionless::create_package() %>%
  add_resource(resource_name = "deployments", data = deployments, schema = here::here("camtrap-dp", "1.0.2", "deployments-table-schema-acoustic.json")) %>%
  add_resource(resource_name = "media", data = media, schema = here::here("camtrap-dp", "1.0.2",  "media-table-schema-acoustic.json")) %>%
  add_resource(resource_name = "observations", data = observations, schema = here::here("camtrap-dp", "1.0.2", "observations-table-schema-acoustic.json")) %>%
  append(c(
    name = "camtrap-dp-acoustic-maryland-dataset", # Required
    profile = here::here("camtrap-dp", "1.0.2", "camtrap-dp-profile-acoustic.json"), # Required
    id = NULL,
    created = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"), # Required
    title = "Maryland: marine mammals detections",
    description = NULL,
    version = "1.0",
    keywords = c("acoustic", "marine mammals", "passive acoustic monitoring"),
    image = NULL,
    homepage = NULL,
    sources = NULL,
    bibliographicCitation = "Bailey H, Rice A, Wingfield JE, Hodge KB, Estabrook BJ, Hawthorne D, Garrod A, Fandel AD, Fouda
        L, McDonald E, Grzyb E, Fletcher W, Hoover AL. 2018. Determining habitat use by marine
        mammals and ambient noise levels using passive acoustic monitoring offshore of Maryland. Sterling
        (VA): US Department of the Interior, Bureau of Ocean Energy Management. OCS Study BOEM
        2019-018. 232 p.",
    coordinatePrecision = NULL,
    spatial = NULL, # Required, but keep this empty, this comes later
    temporal = NULL, # Required, but keep this empty, this comes later
    taxonomic = NULL, # Required, but keep this empty, this comes later
    relatedIdentifiers = NULL,
    references = NULL
  )) %>%
  frictionless::create_package()

acoustic$contributors <- list( # Required
  list(
    title = "Aaron N. Rice",
    email = "arice@cornell.edu",
    path = "https://orcid.org/0000-0002-8598-9705",
    role = "Science director",
    organization = "Bioacoustics Research Program, Cornell Laboratory of Ornithology"
  )
  # list(
  #   title = "",
  #   email = "",
  #   path = "",
  #   role = "",
  #   organization = ""
  # ),
)

acoustic$licenses <- list(
  list(
    name = "CC0-1.0",
    scope = "data" # Required
  ),
  list(
    path = "http://creativecommons.org/licenses/by/4.0/",
    scope = "media" # Required
  )
)

acoustic$project <- list( # Required
  # id = "",
  title = "Maryland", # Required
  # acronym = "",
  description = "Determining Habitat Use by Marine Mammals and Ambient Noise Levels Using Passive Acoustic Monitoring Offshore of Maryland",
  protocolType = list(c("acoustic")), # Required
  samplingDesign = "systematicRandom", # Required
  path = "https://thesoundofnorway.com/",
  captureMethod = list(c("continuous")), # Required | Bug, see https://github.com/frictionlessdata/frictionless-r/issues/276
  classificationEffort = NA_character_,
  individualAnimals = FALSE, # Required
  observationLevel = list(c("media")) # Required | Bug, see https://github.com/frictionlessdata/frictionless-r/issues/276
)
```

## Create taxonomy

```{r}
map_taxonomy <- function(taxonomy) {
  list(
    scientificName = taxonomy$scientificName,
    vernacularNames = list(
      eng = taxonomy$EnglishName
      )
  )
}
```

# Write Data Package

```{r}
frictionless::write_package(acoustic, here::here("datasets", "maryland", "data", "interim"))
```

# Read with camtrapdp

This automatically updates the spatial, temporal and taxonomic coverage in the metadata, based on the data.

```{r}
acoustic_dp <- camtrapdp::read_camtrapdp(here::here("datasets", "maryland", "data", "interim", "datapackage.json"))
```

# Write Camtrap DP

```{r}
camtrapdp::write_camtrapdp(acoustic_dp, here::here("datasets", "maryland", "data", "processed"))
```

# Inspect final tables and compare with raw data
```{r}
# Create inspection tables
inspect_data_raw <- data_raw %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "column", values_to = "n_unique")

inspect_selections <- selections %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "column", values_to = "n_unique")

inspect_deployments <- deployments %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "column", values_to = "n_unique")

inspect_media <- media %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "column", values_to = "n_unique")

inspect_observations <- observations %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "column", values_to = "n_unique")

# Export inspection tables to a single text file
inspection_tables <- list(
  inspect_data_raw = inspect_data_raw, # data_raw
  inspect_selections = inspect_selections, # selections
  inspect_deployments = inspect_deployments, # deployments
  inspect_media = inspect_media, # media
  inspect_observations = inspect_observations  # observations
)

inspect_txt_path <- here::here("datasets", "maryland", "data", "interim", "inspect_tables_concat.txt")
fileConn <- file(inspect_txt_path, open = "wt")
for (tbl_name in names(inspection_tables)) {
  tbl <- inspection_tables[[tbl_name]]
  # Write table name
  writeLines(paste0("## ", tbl_name), fileConn)
  # Write table content
  write.table(tbl, fileConn, sep = "\t", row.names = FALSE, quote = FALSE, col.names = TRUE)
  # Write empty line
  writeLines("", fileConn)
}
close(fileConn)
```

# Vizualise first rows from raw and processed data
```{r}
# Sample 3 random rows from each table
set.seed(123) # for reproducibility

sample_data_raw <- data_raw[sample(nrow(data_raw), min(3, nrow(data_raw))), ]
sample_selections <- selections[sample(nrow(selections), min(3, nrow(selections))), ]
sample_deployments <- deployments[sample(nrow(deployments), min(3, nrow(deployments))), ]
sample_media <- media[sample(nrow(media), min(3, nrow(media))), ]
sample_observations <- observations[sample(nrow(observations), min(3, nrow(observations))), ]

# Create sample tables list
sample_tables <- list(
  sample_data_raw = sample_data_raw, # data_raw
  sample_selections = sample_selections, # selections
  sample_deployments = sample_deployments, # deployments
  sample_media = sample_media, # media
  sample_observations = sample_observations  # observations
)

# Create table structure comparison
library(knitr)
library(kableExtra)

# Function to get table structure info
get_table_structure <- function(df, table_name) {
  structure_info <- data.frame(
    table = table_name,
    column = names(df),
    type = sapply(df, class),
    n_unique = sapply(df, function(x) length(unique(x[!is.na(x)]))),
    n_missing = sapply(df, function(x) sum(is.na(x))),
    example_values = sapply(df, function(x) {
      unique_vals <- unique(x[!is.na(x)])
      if(length(unique_vals) > 0) {
        paste(head(unique_vals, 2), collapse = ", ")
      } else {
        "All NA"
      }
    }),
    stringsAsFactors = FALSE
  )
  return(structure_info)
}

# Get structure for all tables
all_structures <- list()
all_structures[["data_raw"]] <- get_table_structure(data_raw, "data_raw")
all_structures[["selections"]] <- get_table_structure(selections, "selections") 
all_structures[["deployments"]] <- get_table_structure(deployments, "deployments")
all_structures[["media"]] <- get_table_structure(media, "media")
all_structures[["observations"]] <- get_table_structure(observations, "observations")

# Combine all structures
combined_structure <- do.call(rbind, all_structures)

# Display structure comparison
print("=== TABLE STRUCTURE COMPARISON ===")
print(combined_structure)

# Show sample data from each table
print("\n=== SAMPLE DATA FROM EACH TABLE ===")
for (tbl_name in names(sample_tables)) {
  cat("\n## ", toupper(tbl_name), " (", nrow(sample_tables[[tbl_name]]), " rows, ", ncol(sample_tables[[tbl_name]]), " columns) ##\n")
  print(head(sample_tables[[tbl_name]], 2))
}

# Export structure comparison to CSV
structure_csv_path <- here::here("datasets", "maryland", "data", "interim", "table_structure_comparison.csv")
write.csv(combined_structure, structure_csv_path, row.names = FALSE)

# Export to Excel with multiple sheets
library(openxlsx)
excel_path <- here::here("datasets", "maryland", "data", "interim", "table_comparison.xlsx")
wb <- createWorkbook()

# Add sample data sheets for each table
for (tbl_name in names(sample_tables)) {
  # Remove "sample_" prefix from sheet names
  sheet_name <- gsub("^sample_", "", tbl_name)
  # Change selections to selections_raw
  if (sheet_name == "selections") {
    sheet_name <- "selections_raw"
  }
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, sample_tables[[tbl_name]])
}

# Save Excel file
saveWorkbook(wb, excel_path, overwrite = TRUE)

cat("\nFiles exported:")
cat("\n- Structure comparison CSV:", structure_csv_path)
cat("\n- Excel file with all tables:", excel_path)
```