---
title: "acoustic_to_camtrapdp"
author: 
- "Sanne Govaert"
- "Julia Wiel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install packages

Before we begin, we need to ensure that all required R packages are installed. This includes packages for data handling, working with frictionless data, and the development version of `camtrapdp` from GitHub.

```{r}
# Install required packages if not already installed
rlang::check_installed(c(
  "here",
  "dplyr",
  "tidyr",
  "frictionless",
  "lubridate",
  "devtools"
))

# We need the development version of camtrapdp from Github:
devtools::install_github("inbo/camtrapdp")
```

## Load libraries

Next, we load all the necessary libraries.

```{r}
library(here)
library(dplyr)
library(tidyr)
library(frictionless)
library(camtrapdp)
library(lubridate)
library(readxl)
```

# Import data

```{r}
data_raw <- read_xlsx(
  here::here("datasets", "maryland", "data", "raw", "2018_UnivMD_Maryland_86896 DEP1.xlsx"),
  sheet = "DATA TABLE FOR EXPORT"
)
set.seed(123) # Set seed for reproducible sampling
selections_raw <- read.delim2(
  here::here( "datasets", "maryland", "data", "raw", "Fin_merged.selections.txt")
) |>
  sample_n(500)
```

# Inspect data

```{r}
(data_raw_unique_values_per_column <-
  data_raw |>
  dplyr::summarise(
    dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))
  ) |>
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "column", values_to = "n_unique"
  )
)

(selections_unique_values_per_column <-
  selections_raw |>
  dplyr::summarise(
    dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))
  ) |>
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "column", values_to = "n_unique"
  )
)
```

# Prepare data

Deduce ID's and scientificName, and format date-time data.

```{r}
data <- data_raw

# Parse deploymentStart and deploymentEnd in UTC
data <- data |>
  dplyr::mutate(
    # Clean deployTime and recoveryTime
    deployTime_clean = sub("^\\d{4}-\\d{2}-\\d{2} ", "", deployTime),
    recoveryTime_clean = sub("^\\d{4}-\\d{2}-\\d{2} ", "", recoveryTime),
    # Combine date and time for deploymentStart and End
    deploymentStart = lubridate::ymd_hms(paste(deployDate, deployTime_clean)),
    deploymentEnd = lubridate::ymd_hms(paste(recoveryDate, recoveryTime_clean))
  )
```

```{r}
selections <- selections_raw

# Build deploymentID
# Extract 6th element from Begin.Path (split by single backslash)
split_paths <- strsplit(selections$Begin.Path, "\\", fixed = TRUE)
selections$path_element_6 <- sapply(split_paths, function(x) if(length(x) >= 6) x[6] else NA_character_)
selections$deploymentID <- paste(selections$path_element_6, "_Dep01_20180711_M", selections$Channel, sep = "")

## Timestamp of the selection
selections$timestamp <- sapply(sub("\\.aif$", "", selections$Begin.File), function(x) {
  parts <- strsplit(x, "_")[[1]]
  # Assume last two substrings are date and time
  date_str <- parts[length(parts) - 1]
  time_str <- parts[length(parts)]
  # Try to parse date and time
  dt <- suppressWarnings(lubridate::ymd_hms(paste(date_str, time_str), tz = "UTC"))
  # Format as ISO 8601 with offset
  ts <- format(dt, "%Y-%m-%dT%H:%M:%S%z", tz = "UTC")
  ts <- sub("(\\+\\d{2})(\\d{2})$", "\\1:\\2", ts)
  return(ts)
})

## ObservationID
selections$observationID <- paste(
  sub("\\.aif$", "", selections$Begin.File),
  selections$Channel,
  selections$Selection,
  sep = "_"
)

## scientificName from Tags
selections <- selections |>
  dplyr::mutate(
    scientificName = dplyr::case_when(
      Tags == "1-4-f" ~ "Balaenoptera physalus",
      Tags == "1-2-f" ~ "Balaenoptera physalus",
      Tags == "1-1-f" ~ "Balaenoptera physalus",
      TRUE ~ NA_character_
    )
  )

## mediaID
selections$mediaID <- sub("\\.aif$", "", selections$Begin.File)

## event start and end
selections <- selections |>
  dplyr::mutate(
    eventStart = format(
      lubridate::mdy_hms(paste(Date.Check, Time.Check), tz = "UTC"),
      "%Y-%m-%dT%H:%M:%S%z",
      tz = "UTC"
    ) |>
      sub("(\\+\\d{2})(\\d{2})$", "\\1:\\2", x = .),
    # Calculate duration in seconds and add to eventStart
    duration_seconds = as.numeric(End.Time..s.) - as.numeric(Begin.Time..s.),
    eventEnd = format(
      lubridate::mdy_hms(paste(Date.Check, Time.Check), tz = "UTC") + lubridate::seconds(duration_seconds),
      "%Y-%m-%dT%H:%M:%S%z",
      tz = "UTC"
    ) |>
      sub("(\\+\\d{2})(\\d{2})$", "\\1:\\2", x = .)
  ) |>
  dplyr::select(-duration_seconds)

## Generate fixed random eventID for repeatability
set.seed(456) # Fixed seed for repeatability
selections$eventID <- paste0("event_", sprintf("%06d", sample(100000:999999, nrow(selections), replace = FALSE)))
```

# Darwin Core mapping

## Create deployments resource

```{r}
deployments <-
  data |>
  dplyr::mutate(
    deploymentID = paste(KEY_deployment_ID, nominalID, sep = "_"), # Required
    locationID = data$nominalID,
    locationName = NA_character_,
    latitude = data$latitudeDeployed_DecDeg, # Required
    longitude = data$longitudeDeployed_DecDeg, # Required
    coordinateUncertainty = NA_integer_,
    elevation = NA_real_,
    deploymentStart = data$deploymentStart, # Required
    deploymentEnd = data$deploymentEnd, # Required
    setupBy = NA_character_,
    deviceID = data$containerID,
    deviceModel = data$recorderDeviceType,
    devicePlatform = "buoy",
    deviceDelay = NA_integer_,
    deviceHeight = NA_real_,
    deviceDepth = data$deploymentDepth_Meters,
    deviceTilt = NA_integer_,
    deviceHeading = NA_integer_,
    recordingSchedule = "continuous",
    detectionDistance = NA_real_,
    baitUse = NA,
    locationType = NA_character_,
    habitat = NA_character_,
    deploymentGroups = data$arrayName,
    deploymentTags = NA_character_,
    deploymentComments = data$notes,
    .keep = "none"
  ) |>
  dplyr::arrange(deploymentID) |>
  dplyr::select(
    deploymentID, locationID, locationName, latitude, longitude,
    coordinateUncertainty, elevation, deploymentStart, deploymentEnd, setupBy,
    deviceID, deviceModel, devicePlatform, deviceDelay, deviceHeight,
    deviceDepth, deviceTilt, deviceHeading, recordingSchedule,
    detectionDistance, baitUse, locationType, habitat, deploymentGroups,
    deploymentTags, deploymentComments
  )
```

## Create media resource

```{r}
media <-
  selections |>
  dplyr::mutate(
    mediaID = selections$mediaID, # Required
    deploymentID = selections$deploymentID, # Required
    captureMethod = "schedule",
    timestamp = selections$timestamp, # Required
    duration = NA_real_,
    filePath = selections$Begin.Path, # Required
    filePublic = FALSE, # Required
    fileName = NA_character_,
    fileMediatype = "audio/aiff", # Required
    exifData = NA,
    bitDepth = 16,
    samplingFrequency = 16000,
    gain = 23.5,
    channels = NA_integer_,
    favorite = NA,
    mediaComments = NA_character_,
    .keep = "none"
  ) |>
  dplyr::arrange(mediaID) |>
  dplyr::distinct() |>
  dplyr::select(
    mediaID, deploymentID, captureMethod, timestamp, duration, filePath,
    filePublic, fileName, fileMediatype, exifData, bitDepth, samplingFrequency,
    gain, channels, duration, favorite, mediaComments
  )
```

## Create observations resource

```{r}
observations <-
  selections |>
  dplyr::mutate(
    observationID = selections$observationID, # Required
    deploymentID = selections$deploymentID, # Required
    mediaID = selections$mediaID,
    eventID = selections$eventID, # Fixed random IDs for repeatability
    eventStart = selections$eventStart, # Required
    eventEnd = selections$eventEnd, # Required
    observationLevel = "interval", # Required
    observationType = "animal", # Required
    deviceSetupType = NA_character_,
    scientificName = selections$scientificName, # Required
    count = NA_integer_,
    lifeStage = NA_character_,
    sex = NA_character_,
    behavior = NA_character_,
    individualID = NA_character_,
    individualPositionRadius = NA_real_,
    individualPositionAngle = NA_real_,
    individualSpeed = NA_real_,
    bboxX = NA_real_,
    bboxY = NA_real_,
    bboxWidth = NA_real_,
    bboxHeight = NA_real_,
    frequencyLow = NA_real_,
    frequencyHigh = NA_real_,
    classificationMethod = "machine",
    classifiedBy = NA_character_,
    classificationTimestamp = NA_character_,
    classificationProbability = selections$Score,
    classificationConfirmation = NA_character_,
    observationTags = NA_character_,
    observationComments = selections$Notes,
    .keep = "none"
  ) |>
  dplyr::arrange(observationID) |>
  dplyr::select(
    observationID, deploymentID, mediaID, eventID, eventStart,
    eventEnd, observationLevel, observationType, deviceSetupType,
    scientificName, count, lifeStage, sex, behavior, individualID,
    individualPositionRadius, individualPositionAngle, individualSpeed,
    bboxX, bboxY, bboxWidth, bboxHeight, frequencyLow, frequencyHigh,
    classificationMethod, classifiedBy, classificationTimestamp,
    classificationProbability, classificationConfirmation, observationTags,
    observationComments
  )
```

# Create data package

```{r}
acoustic <-
  frictionless::create_package() |>
  add_resource(
    resource_name = "deployments",
    data = deployments,
    schema = here::here( "camtrap-dp", "1.0.2", "deployments-table-schema-acoustic.json")
  ) |>
  add_resource(
    resource_name = "media",
    data = media,
    schema = here::here( "camtrap-dp", "1.0.2", "media-table-schema-acoustic.json")
  ) |>
  add_resource(
    resource_name = "observations",
    data = observations,
    schema = here::here( "camtrap-dp", "1.0.2", "observations-table-schema-acoustic.json")
  ) |>
  append(c(
    name = "camtrap-dp-acoustic-maryland-dataset", # Required
    profile = "https://raw.githubusercontent.com/camera-traps/bioacoustics/main/camtrap-dp/1.0.2/camtrap-dp-profile-acoustic.json", # Required
    id = NULL,
    created = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"), # Required
    title = "Maryland: marine mammals detections",
    description = NULL,
    version = "1.0",
    keywords = c("acoustic", "marine mammals", "passive acoustic monitoring"),
    image = NULL,
    homepage = NULL,
    sources = NULL,
    bibliographicCitation = "Bailey H, Rice A, Wingfield JE, Hodge KB, Estabrook BJ, Hawthorne D, Garrod A, Fandel AD, Fouda L, McDonald E, Grzyb E, Fletcher W, Hoover AL. 2018. Determining habitat use by marine mammals and ambient noise levels using passive acoustic monitoring offshore of Maryland. Sterling (VA): US Department of the Interior, Bureau of Ocean Energy Management. OCS Study BOEM 2019-018. 232 p.",
    coordinatePrecision = NULL,
    spatial = NULL, # Required, but keep this empty, this comes later
    temporal = NULL, # Required, but keep this empty, this comes later
    taxonomic = NULL, # Required, but keep this empty, this comes later
    relatedIdentifiers = NULL,
    references = NULL
  )) |>
  frictionless::create_package()

acoustic$contributors <- list( # Required
  list(
    title = "Aaron N. Rice",
    email = "arice@cornell.edu",
    path = "https://orcid.org/0000-0002-8598-9705",
    role = "principalInvestigator",
    organization = "Bioacoustics Research Program, Cornell Laboratory of Ornithology"
  )
)

acoustic$licenses <- list(
  list(
    name = "CC0-1.0",
    scope = "data" # Required
  ),
  list(
    path = "http://creativecommons.org/licenses/by/4.0/",
    scope = "media" # Required
  )
)

acoustic$project <- list( # Required
  # id = "",
  title = "Maryland", # Required
  # acronym = "",
  description = "Determining Habitat Use by Marine Mammals and Ambient Noise Levels Using Passive Acoustic Monitoring Offshore of Maryland",
  protocolType = list(c("acoustic")), # Required
  samplingDesign = "systematicRandom", # Required
  # path = NA_character_,
  captureMethod = list(c("continuous")), # Required | Bug, see https://github.com/frictionlessdata/frictionless-r/issues/276
  # classificationEffort = NA_character_,
  individualAnimals = FALSE, # Required
  observationLevel = list(c("media")) # Required | Bug, see https://github.com/frictionlessdata/frictionless-r/issues/276
)
```

# Write Data Package

```{r}
frictionless::write_package(acoustic, here::here( "datasets", "maryland", "data", "interim"))
```

# Read with camtrapdp

This automatically updates the spatial, temporal and taxonomic coverage in the metadata, based on the data.

```{r}
acoustic_dp <- camtrapdp::read_camtrapdp(here::here( "datasets", "maryland", "data", "interim", "datapackage.json"))
```

# Write Camtrap DP

```{r}
camtrapdp::write_camtrapdp(acoustic_dp, here::here( "datasets", "maryland", "data", "processed"))
```
