---
title: "acoustic_to_camtrapdp"
author: 
- "Sanne Govaert"
- "Julia Wiel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(here)
library(dplyr)
library(tidyr)
library(frictionless)
library(camtrapdp)
library(lubridate)
library(readxl)
```

# Import data

```{r}
data_raw <- read_xlsx(here::here("data", "raw", "maryland",
 "2018_UnivMD_Maryland_86896 DEP1.xlsx"), sheet = 'DATA TABLE FOR EXPORT')
selections <- read.delim2(here::here("data", "raw", "maryland",
 "Fin_merged.selections.txt")) %>% sample_n(500)
```

# Create resources

## Inspect data

```{r}
(data_raw_unique_values_per_column <-
  data_raw %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "column", values_to = "n_unique"))

(selections_unique_values_per_column <-
  selections %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), ~ dplyr::n_distinct(.))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "column", values_to = "n_unique"))
```

## Add information to dataset

# data
```{r}
# Parse deploymentStart and deploymentEnd in UTC
data <- data_raw %>%
  dplyr::mutate(
    # Clean deployTime and recoveryTime
    deployTime_clean = sub("^\\d{4}-\\d{2}-\\d{2} ", "", deployTime),
    recoveryTime_clean = sub("^\\d{4}-\\d{2}-\\d{2} ", "", recoveryTime),
    # Combine date and time for deployment start/end
    deploy_datetime_str = paste(deployDate, deployTime_clean),
    recovery_datetime_str = paste(recoveryDate, recoveryTime_clean)
  ) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    deploymentStart = format(lubridate::ymd_hms(deploy_datetime_str, tz = "UTC"), "%Y-%m-%dT%H:%M:%S%z", tz = "UTC"),
    deploymentStart = sub("(\\+\\d{2})(\\d{2})$", "\\1:\\2", deploymentStart),
    deploymentEnd = format(lubridate::ymd_hms(recovery_datetime_str, tz = "UTC"), "%Y-%m-%dT%H:%M:%S%z", tz = "UTC"),
    deploymentEnd = sub("(\\+\\d{2})(\\d{2})$", "\\1:\\2", deploymentEnd)
  ) %>%
  dplyr::ungroup()

```

# selections
```{r}
## Create deploymentID from Begin.Path in selections

# Split Begin.Path by "/"
split_path <- strsplit(selections$Begin.Path, '\\\\')
# Extract elements by position (e.g., 5th, 6th, 7th)
deploymentID <- sapply(split_path, function(x) {
  # Split x[7] by "_"
  parts <- strsplit(x[7], "_")[[1]]
  # Remove the first term and paste the rest back together
  new_x7 <- paste(parts[-1], collapse = "_")
  new_x7 <- gsub("MD", "Dep", new_x7)
  # Combine x[6] and the modified x[7]
  paste(x[6], new_x7, sep = "_")
})
# Add to selections
selections$deploymentID <- deploymentID

## Timestamp of the selection
selections$timestamp <- sapply(sub("\\.aif$", "", selections$Begin.File), function(x) {
  parts <- strsplit(x, "_")[[1]]
  # Assume last two substrings are date and time
  date_str <- parts[length(parts)-1]
  time_str <- parts[length(parts)]
  # Try to parse date and time
  dt <- suppressWarnings(lubridate::ymd_hms(paste(date_str, time_str), tz = "UTC"))
  # Format as ISO 8601 with offset
  ts <- format(dt, "%Y-%m-%dT%H:%M:%S%z", tz = "UTC")
  ts <- sub("(\\+\\d{2})(\\d{2})$", "\\1:\\2", ts)
  return(ts)
})

## ObservationID
selections$observationID <- paste(selections$Begin.File, selections$Selection,
 sep = "_")
```
## Create deployments resource

```{r}
deployments <-
  data %>% 
  dplyr::mutate(
    deploymentID = data$KEY_deployment_ID, # Required
    locationID = data$nominalID,
    locationName = NA_character_,
    latitude = data$latitudeDeployed_DecDeg, # Required
    longitude = data$longitudeDeployed_DecDeg, # Required
    coordinateUncertainty = NA_integer_,
    elevation = NA_real_,
    deploymentStart = data$deploymentStart, # Required
    deploymentEnd = data$deploymentEnd, # Required
    setupBy = NA_character_,
    deviceID = data$containerID,
    deviceModel = data$recorderDeviceType,
    devicePlatform = "mooring",
    deviceDelay = NA_integer_,
    deviceHeight = NA_real_,
    deviceDepth = data$deploymentDepth_Meters,
    deviceTilt = NA_integer_,
    deviceHeading = NA_integer_,
    recordingSchedule = "continuous",
    detectionDistance = NA_real_,
    timestampIssues = NA,
    baitUse = NA,
    featureType = NA_character_,
    habitat = NA_character_,
    deploymentGroups = data$arrayName,
    deploymentTags = NA_character_,
    deploymentComments = data$notes,
    .keep = "none"
  ) %>% 
  dplyr::arrange(deploymentID) %>%
  dplyr::select(
    deploymentID, locationID, locationName, latitude, longitude,
    coordinateUncertainty, elevation, deploymentStart, deploymentEnd, setupBy,
    deviceID, deviceModel, devicePlatform, deviceDelay, deviceHeight,
    deviceDepth, deviceTilt, deviceHeading, recordingSchedule,
    detectionDistance, timestampIssues, baitUse, featureType, habitat,
    deploymentGroups, deploymentTags, deploymentComments
  )
```

## Create media resource

```{r}
media <-
  selections %>% 
  dplyr::mutate(
    mediaID = selections$Begin.File, # Required
    deploymentID = selections$deploymentID, # Required
    captureMethod = "schedule",
    timestamp = selections$timestamp, # Required 
    filePath = selections$Begin.Path,# Required
    filePublic = FALSE, # Required
    fileName = NA_character_,
    fileMediatype = "audio/aiff", # Required
    samplingFrequency = 16000,
    bitDepth = 16,
    gain = 23.5,
    channels = NA_integer_,
    duration = NA_real_,
    exifData = NA,
    favorite = NA,
    mediaComments = NA_character_,
    .keep = "none"
  ) %>% 
  dplyr::arrange(mediaID) %>%
  dplyr::distinct() %>% 
  dplyr::select(
    mediaID, deploymentID, captureMethod, timestamp, filePath, filePublic,
    fileName, fileMediatype, samplingFrequency, bitDepth, gain, channels,
    duration, exifData, favorite, mediaComments
  )
```

## Create observations resource

```{r}
observations <-
  selections %>% 
  dplyr::mutate(
    observationID = selections$observationID, # Required
    deploymentID = selections$deploymentID, # Required
    mediaID = selections$Begin.File,
    eventID = NA_character_, # Necessary to use read_camtrapdp(), see https://github.com/inbo/camtrapdp/issues/162
    eventStart = selections$Begin.Time..s., # format(data$detected_time_parsed, "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"), # Required
    eventEnd = selections$End.Time..s., # Required
    observationLevel = "interval", # Required
    observationType = "animal", # Required 
    deviceSetupType = NA_character_,
    scientificName = NA_character_,  # Required
    count = NA_integer_,
    lifeStage = NA_character_,
    sex = NA_character_,
    behavior = NA_character_,
    individualID = NA_character_,
    individualPositionRadius = NA_real_,
    individualPositionAngle = NA_real_,
    individualSpeed = NA_real_,
    bboxX = NA_real_,
    bboxY = NA_real_,
    bboxWidth = NA_real_,
    bboxHeight = NA_real_,
    freqLow = NA_real_,
    freqHigh = NA_real_,
    classificationMethod = "machine",
    classifiedBy = NA_character_,
    classificationTimestamp = NA_character_,
    classificationProbability = selections$Score,
    observationTags = selections$Tags,
    observationComments = selections$Notes,
    .keep = "none"
  ) %>% 
  dplyr::arrange(observationID) %>% 
  dplyr::select(
    observationID, deploymentID, mediaID, eventID, eventStart, eventEnd,
    observationLevel, observationType, deviceSetupType, scientificName, count,
    lifeStage, sex, behavior, individualID, individualPositionRadius,
    individualPositionAngle, individualSpeed, bboxX, bboxY, bboxWidth,
    bboxHeight, freqLow, freqHigh, classificationMethod, classifiedBy,
    classificationTimestamp, classificationProbability, observationTags,
    observationComments
  )
```

# Create data package

```{r}
acoustic <-
  frictionless::create_package() %>%
  add_resource(resource_name = "deployments", data = deployments, schema = here::here("camtrap-dp", "deployments-table-schema-acoustic.json")) %>%
  add_resource(resource_name = "media", data = media, schema = here::here("camtrap-dp", "media-table-schema-acoustic.json")) %>%
  add_resource(resource_name = "observations", data = observations, schema = here::here("camtrap-dp", "observations-table-schema-acoustic.json")) %>%
  append(c(
    name = "camtrap-dp-acoustic-maryland-dataset", # Required
    profile = "https://raw.githubusercontent.com/tdwg/camtrap-dp/1.0.1/camtrap-dp-profile.json", # Required (incorrect, hack to use read_camtrapdp())
    id = NULL,
    created = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"), # Required
    title = "Maryland: marine mammals detections",
    description = NULL,
    version = "1.0",
    keywords = c("acoustic", "marine mammals", "passive acoustic monitoring"),
    image = NULL,
    homepage = NULL,
    sources = NULL,
    bibliographicCitation = "Bailey H, Rice A, Wingfield JE, Hodge KB, Estabrook BJ, Hawthorne D, Garrod A, Fandel AD, Fouda
        L, McDonald E, Grzyb E, Fletcher W, Hoover AL. 2018. Determining habitat use by marine
        mammals and ambient noise levels using passive acoustic monitoring offshore of Maryland. Sterling
        (VA): US Department of the Interior, Bureau of Ocean Energy Management. OCS Study BOEM
        2019-018. 232 p.",
    coordinatePrecision = NULL,
    spatial = NULL, # Required, but keep this empty, this comes later
    temporal = NULL, # Required, but keep this empty, this comes later
    taxonomic = NULL, # Required, but keep this empty, this comes later
    relatedIdentifiers = NULL,
    references = NULL
  )) %>%
  frictionless::create_package()

acoustic$contributors <- list( # Required
  list(
    title = "Aaron N. Rice",
    email = "arice@cornell.edu",
    path = "https://orcid.org/0000-0002-8598-9705",
    role = "Science director",
    organization = "Bioacoustics Research Program, Cornell Laboratory of Ornithology"
  )
  # list(
  #   title = "",
  #   email = "",
  #   path = "",
  #   role = "",
  #   organization = ""
  # ),
)

acoustic$licenses <- list(
  list(
    name = "CC0-1.0",
    scope = "data" # Required
  ),
  list(
    path = "http://creativecommons.org/licenses/by/4.0/",
    scope = "media" # Required
  )
)

acoustic$project <- list( # Required
  # id = "",
  title = "Maryland", # Required
  # acronym = "",
  description = "Determining Habitat Use by Marine Mammals and Ambient Noise Levels Using Passive Acoustic Monitoring Offshore of Maryland",
  protocolType = list(c("acoustic")), # Required
  samplingDesign = "systematicRandom", # Required
  path = "https://thesoundofnorway.com/",
  captureMethod = list(c("continuous")), # Required | Bug, see https://github.com/frictionlessdata/frictionless-r/issues/276
  classificationEffort = NA_character_,
  individualAnimals = FALSE, # Required
  observationLevel = list(c("media")) # Required | Bug, see https://github.com/frictionlessdata/frictionless-r/issues/276
)
```

## Create taxonomy

```{r}
map_taxonomy <- function(taxonomy) {
  list(
    scientificName = taxonomy$scientificName,
    vernacularNames = list(
      eng = taxonomy$EnglishName
      )
  )
}
```

```{r}
#TODO Add Taxonomy
# taxonomy_list <- apply(taxonomy, 1, as.list)
# acoustic$taxonomic <- purrr::map(taxonomy_list, map_taxonomy)
```

# Write Data Package

```{r}
frictionless::write_package(acoustic, here::here("data", "interim", "maryland"))
```

# Read with camtrapdp

This automatically updates the spatial, temporal and taxonomic coverage in the metadata, based on the data.

```{r}
acoustic_dp <- camtrapdp::read_camtrapdp(here::here("data", "interim", "maryland", "datapackage.json"))
```

# Write Camtrap DP

```{r}
camtrapdp::write_camtrapdp(acoustic_dp, here::here("data", "processed", "maryland"))
```
