---
title: "acoustic_to_camtrapdp"
author: 
- "Sanne Govaert"
- "Julia Wiel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r}
library(here)
library(dplyr)
library(tidyr)
library(frictionless)
library(camtrapdp)
library(lubridate)
```

# Import data

```{r}
# Read main acoustic observations data
data_raw <- read.csv(here::here("datasets", "sound_of_norway", "data", "raw", "sound_of_norway_subsample.csv"))
# Read taxonomy mapping (scientific and English names)
taxonomy <- read.delim2(
  file = here::here("datasets", "sound_of_norway", "data", "raw", "labels_en.txt"),
  sep = "_",
  header = FALSE,
  col.names = c("scientificName", "EnglishName")
)
```

## Inspect data

```{r}
# Summarize number of unique values per column in the raw data
data_raw %>%
  dplyr::summarise(dplyr::across(dplyr::everything(), dplyr::n_distinct)) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "column", values_to = "n_unique")
```

## Add information to dataset

```{r}
# Add new columns to the data for eventEnd, timestamp and formatting
data <-
  data_raw %>% 
  dplyr::mutate(
    # Parse detected time to datetime object in UTC
    detected_time_parsed = lubridate::ymd_hms(.data$detected_time, tz = "UTC"),
    # Calculate eventEnd
    eventEnd = .data$detected_time_parsed + .data$end_secs,
    # Format event end time to ISO 8601 string
    eventEnd_iso = format(
      .data$eventEnd, format = "%Y-%m-%dT%H:%M:%OSZ", tz = "UTC"
      ),
    # Extract interim timestamp from audio file name
    timestamp_interim = sub(".mp3", "", basename(.data$audio_link)),
    # Format timestamp
    timestamp = gsub("_", ":", .data$timestamp_interim) # format %Y-%m-%dT%H:%M:%S.%f%z
  )

# Add scientificName to data by joining by EnglishName (tags)
data <- dplyr::left_join(data, taxonomy, dplyr::join_by(tags == EnglishName))

# Extract deployment-level information
deployments_raw <- data %>%
  group_by(site) %>%
  dplyr::mutate(
    # deploymentStart and deploymentEnd based on first and last detection for each site
    deploymentStart = min(detected_time_parsed),
    deploymentEnd = max(detected_time_parsed),
    deploymentID = .data$site,
    ) %>%
  dplyr::select(site, latitude, longitude, deploymentStart, deploymentEnd, 
                deploymentID, recorder) %>% 
  dplyr::distinct() %>% 
  dplyr::ungroup()
```

# Create resources

## Create deployments resource

```{r}
deployments <-
  deployments_raw %>% 
  dplyr::mutate(
    deploymentID = .data$deploymentID, # Required
    locationID = NA_character_,
    locationName = .data$site,
    latitude = .data$latitude, # Required
    longitude = .data$longitude, # Required
    coordinateUncertainty = NA_integer_,
    elevation = NA_real_,
    deploymentStart = .data$deploymentStart, # Required
    deploymentEnd = .data$deploymentEnd, # Required
    setupBy = NA_character_,
    deviceID = .data$recorder,
    deviceModel = "Bugg v2022",
    devicePlatform = "pole",
    deviceDelay = NA_integer_,
    deviceHeight = 1.5,
    deviceDepth = NA_real_,
    deviceTilt = NA_integer_,
    deviceHeading = NA_integer_,
    recordingSchedule = "continuous",
    detectionDistance = NA_real_,
    baitUse = NA,
    locationType = NA_character_,
    habitat = NA_character_,
    deploymentGroups = NA_character_,
    deploymentTags = NA_character_,
    deploymentComments = NA_character_,
    .keep = "none"
  ) %>% 
  dplyr::arrange(deploymentID) %>% 
  dplyr::select(
    deploymentID, locationID, locationName, latitude, longitude, 
    coordinateUncertainty, elevation, deploymentStart, deploymentEnd, setupBy, 
    deviceID, deviceModel, devicePlatform, deviceDelay, deviceHeight, 
    deviceDepth, deviceTilt, deviceHeading, recordingSchedule, detectionDistance,
    baitUse, locationType, habitat, deploymentGroups, deploymentTags,
    deploymentComments
  )
```

## Create media resource

```{r}
media <-
  data %>% 
  dplyr::mutate(
    mediaID = .data$audio_id, # Required
    deploymentID = .data$site, # Required
    captureMethod = "recordingSchedule",
    timestamp = .data$timestamp, # Required 
    duration = 300,
    filePath = .data$audio_link,# Required
    filePublic = FALSE, # Required
    fileName = NA_character_,
    fileMediatype = "audio/mpeg", # Required
    exifData = NA,
    bitDepth = 16,
    samplingFrequency = 44100,
    gain = NA_real_,
    channels = 1,
    favorite = NA,
    mediaComments = NA_character_,
    .keep = "none"
  ) %>% 
  dplyr::arrange(mediaID) %>%
  dplyr::distinct() %>% 
  dplyr::select(
    mediaID, deploymentID, captureMethod, timestamp, duration, filePath, 
    filePublic, fileName, fileMediatype, exifData, bitDepth, 
    samplingFrequency, gain, channels, duration, favorite, mediaComments
  )
```

## Create observations resource

```{r}
observations <-
  data %>% 
  dplyr::mutate(
    observationID = .data$id, # Required
    deploymentID = .data$site, # Required
    mediaID = .data$audio_id,
    eventID = NA_character_, # Necessary to use read_camtrapdp(), see https://github.com/inbo/camtrapdp/issues/162
    eventStart = .data$detected_time, # format(.data$detected_time_parsed, "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"), # Required
    eventEnd = .data$eventEnd_iso, # Required
    observationLevel = "interval", # Required
    observationType = "animal", # Required 
    deviceSetupType = NA_character_,
    scientificName = .data$scientificName,  # Required
    count = NA_integer_,
    lifeStage = NA_character_,
    sex = NA_character_,
    behavior = NA_character_,
    individualID = NA_character_,
    individualPositionRadius = NA_real_,
    individualPositionAngle = NA_real_,
    individualSpeed = NA_real_,
    bboxX = NA_real_,
    bboxY = NA_real_,
    bboxWidth = NA_real_,
    bboxHeight = NA_real_,
    frequencyLow = NA_real_,
    frequencyHigh = NA_real_,
    classificationMethod = "machine",
    classifiedBy = .data$analysis,
    classificationTimestamp = "2023-06-23T12:23:00Z",
    classificationProbability = .data$confidence,
    classificationConfirmation = NA_character_,
    observationTags = .data$tags,
    observationComments = NA_character_,
    .keep = "none"
  ) %>% 
  dplyr::arrange(observationID) %>% 
  dplyr::select(
    observationID, deploymentID, mediaID, eventID, eventStart,
    eventEnd, observationLevel, observationType, deviceSetupType,
    scientificName, count, lifeStage, sex, behavior, individualID,
    individualPositionRadius, individualPositionAngle, individualSpeed,
    bboxX, bboxY, bboxWidth, bboxHeight, frequencyLow, frequencyHigh,
    classificationMethod, classifiedBy, classificationTimestamp,
    classificationProbability, classificationConfirmation, observationTags,
    observationComments
  )
```

# Create data package

```{r}
# Build frictionless data package with each resource and metadata
acoustic <-
  frictionless::create_package() %>%
  add_resource(resource_name = "deployments", data = deployments, schema = here::here("camtrap-dp", "1.0.2", "deployments-table-schema-acoustic.json")) %>%
  add_resource(resource_name = "media", data = media, schema = here::here("camtrap-dp", "1.0.2", "media-table-schema-acoustic.json")) %>%
  add_resource(resource_name = "observations", data = observations, schema = here::here("camtrap-dp", "1.0.2", "observations-table-schema-acoustic.json")) %>%
  append(c(
    name = "sound-of-norway", # Required
    profile = "https://raw.githubusercontent.com/camera-traps/bioacoustics/main/camtrap-dp/1.0.2/camtrap-dp-profile-acoustic.json", # Required
    id = NULL,
    created = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"), # Required
    title = "The Sound of Norway: bird detections",
    description = NULL,
    version = "1.0",
    keywords = c("acoustic", "birds", "passive acoustic monitoring"),
    image = NULL,
    homepage = NULL,
    sources = NULL,
    bibliographicCitation = "Bick, I. Avery, Vegar Bakkestuen, Benjamin Cretois, Ben Hillier, John A. Kålås, Marius Pedersen, Kiran Raja, et al. ‘National-Scale Acoustic Monitoring of Avian Biodiversity and Phenology’. Ecology, 24 May 2024. https://doi.org/10.1101/2024.05.21.595242",
    coordinatePrecision = NULL,
    spatial = NULL, # Required, but keep this empty, this comes later
    temporal = NULL, # Required, but keep this empty, this comes later
    taxonomic = NULL, # Required, but keep this empty, this comes later
    relatedIdentifiers = NULL,
    references = NULL
  )) %>%
  frictionless::create_package()

# Add contributors metadata to data package
acoustic$contributors <- list( # Required
  list(
    title = "Julia Wiel",
    email = "julia.wiel@nina.no",
    path = "https://orcid.org/0009-0009-7224-3953",
    role = "contributor",
    organization = "Norwegian Institute for Nature Research (NINA)"
  ),
  list(
    title = "Benjamin Cretois",
    email = "benjamin.cretois@nina.no",
    path = "https://orcid.org/0000-0001-8668-3321",
    role = "rightsHolder",
    organization = "Norwegian Institute for Nature Research (NINA)"
  ),
  list(
    title = "Carolyn Rosten",
    email = "carolyn.rosten@nina.no",
    path = "https://orcid.org/0000-0002-1117-444X",
    role = "rightsHolder",
    organization = "Norwegian Institute for Nature Research (NINA)"
  ),
  list(
    title = "Sarab Sethi",
    email = "sarab.sethi@imperial.ac.uk",
    path = "https://orcid.org/0000-0002-5939-0432",
    role = "principalInvestigator",
    organization = "Imperial College London"
  ),
  list(
    title = "I. Avery Bick",
    email = "i.avery.bick@gmail.com",
    path = "https://orcid.org/0000-0002-8007-9649",
    role = "principalInvestigator",
    organization = "Norwegian Institute for Nature Research (NINA)"
  )
)

# Add license metadata for data and media files
acoustic$licenses <- list(
  list(
    name = "CC0-1.0",
    scope = "data" # Required
  ),
  list(
    path = "http://creativecommons.org/licenses/by/4.0/",
    scope = "media" # Required
  )
)

# Add project metadata describing the overarching study design and protocol
acoustic$project <- list( # Required
  # id = "",
  title = "The Sound of Norway", # Required
  # acronym = "",
  description = "The Sound of Norway project monitors ecosystems across the nation with real- time recording devices. Using machine learning to automatically classify bird vocalisations, we can simultaneously track the distribution and migration patterns of many species on vast scales and over long time-periods.",
  protocolType = list(c("acoustic")), # Required
  samplingDesign = "systematicRandom", # Required
  path = "https://thesoundofnorway.com/",
  captureMethod = list(c("continuous")), # Required | Bug, see https://github.com/frictionlessdata/frictionless-r/issues/276
  individualAnimals = FALSE, # Required
  observationLevel = list(c("media")) # Required | Bug, see https://github.com/frictionlessdata/frictionless-r/issues/276
)
```

## Create taxonomy

```{r}
# Function to convert taxonomy data into list format for metadata
map_taxonomy <- function(taxonomy) {
  list(
    scientificName = taxonomy$scientificName,
    vernacularNames = list(
      eng = taxonomy$EnglishName
      )
  )
}
```

```{r}
# Apply mapping function to each row in taxonomy for package metadata
taxonomy_list <- apply(taxonomy, 1, as.list)
acoustic$taxonomic <- purrr::map(taxonomy_list, map_taxonomy)
```

# Write Data Package

## Create frictionless data package

```{r}
# Save frictionless data package (including resources and metadata) to interim directory
frictionless::write_package(acoustic, here::here("datasets", "sound_of_norway", "data", "interim"))
```

## Read with camtrapdp

This automatically updates the spatial, temporal and taxonomic coverage in the metadata, based on the data.

```{r}
# Read camtrap data package to update metadata coverage fields
acoustic_dp <- camtrapdp::read_camtrapdp(here::here("datasets", "sound_of_norway", "data", "interim", "datapackage.json"))
```

## Write Camtrap DP

```{r}
# Write final Camera Trap Data Package to processed directory
camtrapdp::write_camtrapdp(acoustic_dp, here::here("datasets", "sound_of_norway", "data", "processed"))
```
